Detection of an Intrinsic Color Parameter within the Type~Ia Supernovae of the Nearby Supernova Factory

### 2016-09-10 forgerard.tex
forging on

\begin{abstract}
An empirical model for SN~Ia peak magnitudes with two color parameters and dependence on the equivalent widths of CaII, SiII, and SiII velocity
is applied to the supernova sample of the Nearby Supernova Factory.  The peak magnitudes and their colors are found to be 
dependent on the spectral equivalent widths, and two independent color parameters with better than 0.999 confidence.
One parameter, interpreted as due to extrinsic host-galaxy
dust extinction, is consistent with an $R_V=2.5$  \citet{1999PASP..111...63F} dust model.  The second parameter is inconsistent with dust and
is inferred to have origins intrinsic to supernovae.  Our model explains features of external data: 
\textcolor{red}{
the wide range of inferred $R_V$ among
SNe~Ia, the systematically lower $R_V$ of redder supernovae relative to the upper bound of the full population,} and the colors of SN~2014J.
\end{abstract}

@@

We assume 
that  peak intrinsic $UBVRI$ magnitudes are linearly dependent
on the equivalent widths of the CaII H\&K and SiII$\lambda$4141 spectral features
$EW_{Ca}$, $EW_{Si}$,
the SiII$\lambda6355$ position $\lambda_{Si}$ (so also linear with $v_{Si}$):
these spectral features are associated with SN~Ia  spectroscopic diversity  
\citep{2006PASP..118..560B,2009A&A...500L..17B, 2009PASP..121..238B, 2009ApJ...699L.139W, 2011ApJ...729...55F}.
$EW_{Si}$ at peak brightness is an effective proxy for the light-curve shape parameter
\citep{2011A&A...529L...4C}. 
An intrinsic-color parameter $D$ describes  intrinsic magnitude/color variation unaccounted for by the spectral features.
Residual dispersion is described by a Normal distribution with a parameterized covariance matrix
$C_c$.  A grey magnitude offset $\Delta$ is included for each supernova
to capture intrinsic dispersion and peculiar-velocity errors introduced when converting fluxes to luminosity.
The observed magnitudes are linearly dependent on an
extrinsic-color parameter $k$.  The observables
$U_o, B_o, V_o, R_o, I_o$, $EW_{Ca,o}$, $EW_{Si,o}$, $v_{Si,o}$
shown in Table~\ref{data:tab} have Gaussian measurement uncertainty with covariance $C$.
The model is written as
\begin{equation}
\begin{pmatrix}
U\\B\\V\\R\\I
\end{pmatrix}
\sim \mathcal{N}
\left(
\Delta +
\begin{pmatrix}
c_0+\alpha_0 EW_{Ca} + \beta_0 EW_{Si} + \eta_0 \lambda_{Si} + \delta_0 D\\
c_1+\alpha_1 EW_{Ca} + \beta_1 EW_{Si} + \eta_1 \lambda_{Si} + \delta_1 D \\
c_2+\alpha_2 EW_{Ca} + \beta_2 EW_{Si} + \eta_2 \lambda_{Si} + \delta_2 D\\
c_3+\alpha_3 EW_{Ca} + \beta_3 EW_{Si} + \eta_3 \lambda_{Si} + \delta_3 D\\
c_4+\alpha_4 EW_{Ca} + \beta_4 EW_{Si}+ \eta_4 \lambda_{Si} + \delta_4 D
\end{pmatrix}
,C_{c}
\right)
\label{ewsiv:eqn}
\end{equation}
\begin{equation}
\begin{pmatrix}
U_o\\B_o\\ V_o\\R_o\\I_o\\EW_{Si, o}\\ EW_{Ca, o} \\ v_{Si, o}
\end{pmatrix}
\sim \mathcal{N}
\left(
\begin{pmatrix}
U +\gamma_0 k \\B +\gamma_1 k \\V+\gamma_2 k\\R+\gamma_3 k\\I+\gamma_4 k\\
EW_{Si}\\ EW_{Ca} \\ \lambda_{Si}
\end{pmatrix}
,C
\right).
\label{dust:eqn}
\end{equation}
The global parameters $c$, $\alpha$, $\beta$, $\eta$, and $\delta$  are the intercepts and slopes of the linear relationships that
relate intrinsic magnitudes with the spectral-feature parameters and intrinsic-color parameter.
(In this article we freely interchange the parameter indices with  $01234$ or $UBVRI$.)
The global parameters $\gamma$ are the slopes that connect the extrinsic-color
parameter and observed magnitudes.  To constrain the degrees of freedom and degeneracies inherent in the model we impose
\begin{equation}
\langle \Delta \rangle=0, \langle k \rangle=0, \langle D \rangle=0, \gamma_0 > 0, \delta_0 < 0.
\end{equation}
The first two specify the definition of zero color relative to which the color excess is measured.    The latter two exclude degenerate posterior space
associated
with  simultaneous sign flips of
$\delta$--$D$ and $\gamma$--$k$.
Although not formally a degeneracy in the model, practically the parameter pairs  $\delta$--$D$ and $\gamma$--$k$ are degenerate with
each other,
as confirmed by recalculation of the posterior after exchange of the initial  and boundary conditions of $\delta$ and $\gamma$.
As the model cannot distinguish which effect is intrinsic or extrinsic, we do so
by assigning the parameters whose values are consistent with dust to $\gamma$ and the others to $\delta$.

@@

The best-fit distributions for $R_X=\frac{\gamma_X}{\gamma_1-\gamma_2}$ are shown in Figure~\ref{rx:fig}.  
Our measurements of
$R_X$ and $\frac{\gamma_X}{\gamma_2}$, placed at the effective wavelength of the supernova flux that passes
through the synthetic bands, are shown in
Figure~\ref{ccm:fig}.
Also plotted are the expectations of the \citet{1999PASP..111...63F} dust model for different values of its parameter $R^F_V$.
The $R_X$ are consistent with  $R^F_V=2.5$.  The 
 alternative parameterization
 of $\gamma_X/\gamma_2$ has significantly smaller
uncertainties and is also consistent with $R^F_V=2.5$.

@@

\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[width=2.8in]{output11/ccm.pdf}
      \includegraphics[width=2.8in]{output11/ccm2.pdf} 
   \caption{68\% credible intervals of Left: $R_X=\frac{\gamma_X}{\gamma_1-\gamma_2}$; Right: $\frac{\gamma_X}{\gamma_2}$. 
   \citet{1999PASP..111...63F} dust
   predictions for these parameters are overlaid for different values of the dust-parameter $R^F_V$.
   \label{ccm:fig}}
\end{figure}

@@

@@

In the fit to our model, the observed color excess is attributed to   extrinsic
$E_\gamma(B-V)=  0.86 \pm   0.08$ and intrinsic $E_\delta(B-V)=  0.35 \pm   0.10$ contributions.
Among the supernovae in the SNfactory  set used to determine $\gamma$ and $\delta$, the highest median value for 
$E_\gamma(B-V)$ is $  0.33 \pm 0.04$ and for $E_\delta(B-V)$ is
$  0.07 \pm 0.03$ 
(see Figure~\ref{ebv:fig}).
In both parameters SN~2014J lives well outside the range of supernovae used to train the coefficients of the model.

\section{Discussion}
\label{discussion:sec}
We model SNe~Ia broadband optical peak magnitudes allowing for correlations with spectral features at peak and distinct intrinsic and
extrinsic color parameters.  Analyzing SNfactory data with the model, we find significant evidence that the above parameters do
affect supernova magnitudes and colors.  The external color parameter has behavior consistent with $R^F_V=2.5$ \citet{1999PASP..111...63F} dust,
whereas the internal color parameter is inconsistent with dust or a blackbody.  This model  does an excellent job in
describing SN~2014J, a supernova not used in the training that has atypical colors.

\citet{2014ApJ...789...32B, 2015MNRAS.453.3300A} deduce a wide range of dust behavior $1.5<R^F_V<3$ encountered by the SN~Ia population.
\citet{2011ApJ...731..120M, 2011ApJ...729...55F} and the above authors
find that supernovae with large $E_o(B-V)$ 
preferentially exhibit low $R^F_V$. 
\color{red}
In our model internal and external effects for each supernova  conflate into an effective color excess
$E_{eff}(X-V) = E_\gamma(X-V) + E_\delta(X-V)$ and effective $V$ extinction $A_{eff,V} = \gamma_2 k  + \delta_2 D$ 
Recall that the colors where $E_\gamma(B-V)=0$,  $E_\delta(B-V)=0$ are free and arbitrarily
fixed using the  boundary conditions $\langle \Delta \rangle = 0$, $\langle k \rangle = 0$:
there is no unique value for the statistic $R_{eff,V} = A_{eff,V} /E_{eff}(B-V)$, as it depends on the choice of  boundary condition.
For specificity in the current qualitative discussion, 
let us consider a set of SNe~Ia  with parameters set to the median fit values of the SNfactory sample.
We specify   $E_{eff}(U-V)$  and $E_{eff}(B-V)$ by
assigning the blue-extreme colors of this set close to  zero color excess, 
as shown in bottom-left corner in the left plot of Figure~\ref{rveff:fig}.
For this choice of color zeropoints,
$R_{eff,U} - R_{eff,V} = E_{eff}(U-V) /E_{eff}(B-V)$ is plotted
in the right of Figure~\ref{rveff:fig}.
At extreme blue effective color there is a tight range of  $R_{eff,U} - R_{eff,V}$ values.  The distribution spreads to significantly 
lower   $R_{eff,U} - R_{eff,V}$ as supernovae appear redder but
then reconverge to higher, tighter values at extreme red effective color.
The variation in  $R_{eff,U} - R_{eff,V}$ is attributed to the varying contributions of internal and external color excess: the bottom plot of
Figure~\ref{rveff:fig} shows that the shift to lower values occurs when the effective color excess is dominated by the internal contribution. 
Since the standard deviation of the $E_\delta(B-V)$ distribution is  narrow compared to that of
 $E_\gamma(B-V)$, low values of $R_{eff,U} - R_{eff,V}$ can only occur
at low $E_{eff}(X-V)$.
 The  \citet{1999PASP..111...63F} dust model interprets
 low values of $R_{eff,U} - R_{eff,V}$ as implying high values of $R_V$ (see Figure~\ref{ebv:fig}),   and would hence infer a
 broad range of  $R_V$ at bluer colors, and a narrow range of low $R_V$  for the reddest supernovae.

\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[width=2.8in]{output11/rveff.pdf}
   \includegraphics[width=2.8in]{output11/rveff2.pdf}
   \includegraphics[width=2.8in]{output11/rveff3.pdf}
      \caption{\textcolor{red}{Left: Effective color excess $E_{eff}(B-V)$ versus $E_{eff}(U-V)$
   where $E_{eff}(X-V)   = E_\gamma(X-V) + E_\delta(X-V) - z_X$
   for a set of SNe with parameters set to those of the medians of the objects in our analysis.
   The $z_X$ represents a choice of colors at zero color-excess so that the lower left corner points have values approaching zero. 
Right:  The inferred $R_{eff,U}- R_{eff,V} = E_{eff}(U-V)/E_{eff}(B-V)$.
Bottom:  $R_{eff,U}- R_{eff,V}$ as a function of the relative contributions of extrinsic and intrinsic color effects $E_\gamma(B-V)/E_\delta(B-V)$.}
   \label{rveff:fig}}
\end{figure}
\color{black}

>>>>

\begin{abstract}
An empirical model for SN~Ia peak magnitudes with two color parameters and dependence on the equivalent widths of CaII, SiII, and SiII velocity
is applied to the supernova sample of the Nearby Supernova Factory.  The peak magnitudes and their colors are found to be 
dependent on the spectral equivalent widths, and two independent color parameters with better than 0.999 confidence.
One parameter, interpreted as due to extrinsic host-galaxy
dust extinction, is consistent with an $R_V=3.06$  \citet{1999PASP..111...63F} dust model.  The second parameter is inconsistent with dust and
is inferred to have origins intrinsic to supernovae.  Our model explains features of external data: 
\textcolor{red}{
the wide range of inferred $R_V$ among
SNe~Ia, the systematically lower $R_V$ of redder supernovae relative to the upper bound of the full population,} and the colors of SN~2014J.
\end{abstract}

@@

We assume 
that  peak intrinsic $UBVRI$ magnitudes are linearly dependent
on the equivalent widths of the CaII H\&K and SiII$\lambda$4141 spectral features
$EW_{Ca}$, $EW_{Si}$,
the SiII$\lambda6355$ position $\lambda_{Si}$ (so also linear with $v_{Si}$):
these spectral features are associated with SN~Ia  spectroscopic diversity  
\citep{2006PASP..118..560B,2009A&A...500L..17B, 2009PASP..121..238B, 2009ApJ...699L.139W, 2011ApJ...729...55F}.
$EW_{Si}$ at peak brightness is an effective proxy for the light-curve shape parameter
\citep{2011A&A...529L...4C}. 
An intrinsic-color parameter $D$ describes  intrinsic magnitude/color variation unaccounted for by the spectral features.
Residual dispersion is described by a Normal distribution with a parameterized covariance matrix
$C_c$.  A grey magnitude offset $\Delta$ is included for each supernova
to capture intrinsic dispersion and peculiar-velocity errors introduced when converting fluxes to luminosity.
The observed magnitudes are linearly dependent on an
extrinsic-color parameter $k$.  The observables
$U_o, B_o, V_o, R_o, I_o$, $EW_{Ca,o}$, $EW_{Si,o}$, $v_{Si,o}$
shown in Table~\ref{data:tab} have Gaussian measurement uncertainty with covariance $C$.
The model is written as
\begin{equation}
\begin{pmatrix}
U\\B\\V\\R\\I
\end{pmatrix}
\sim \mathcal{N}
\left(
\Delta +
\begin{pmatrix}
c_0+\alpha_0 EW_{Ca} + \beta_0 EW_{Si} + \eta_0 \lambda_{Si} + \delta_0 D\\
c_1+\alpha_1 EW_{Ca} + \beta_1 EW_{Si} + \eta_1 \lambda_{Si} + \delta_1 D \\
c_2+\alpha_2 EW_{Ca} + \beta_2 EW_{Si} + \eta_2 \lambda_{Si} + \delta_2 D\\
c_3+\alpha_3 EW_{Ca} + \beta_3 EW_{Si} + \eta_3 \lambda_{Si} + \delta_3 D\\
c_4+\alpha_4 EW_{Ca} + \beta_4 EW_{Si}+ \eta_4 \lambda_{Si} + \delta_4 D
\end{pmatrix}
,C_{c}
\right)
\label{ewsiv:eqn}
\end{equation}
\begin{equation}
\begin{pmatrix}
U_o\\B_o\\ V_o\\R_o\\I_o\\EW_{Si, o}\\ EW_{Ca, o} \\ v_{Si, o}
\end{pmatrix}
\sim \mathcal{N}
\left(
\begin{pmatrix}
U +\gamma_0 k \\B +\gamma_1 k \\V+\gamma_2 k\\R+\gamma_3 k\\I+\gamma_4 k\\
EW_{Si}\\ EW_{Ca} \\ \lambda_{Si}
\end{pmatrix}
,C
\right).
\label{dust:eqn}
\end{equation}
The global parameters $c$, $\alpha$, $\beta$, $\eta$, and $\delta$  are the intercepts and slopes of the linear relationships that
relate intrinsic magnitudes with the spectral-feature parameters and intrinsic-color parameter.
(In this article we freely interchange the parameter indices with  $01234$ or $UBVRI$.)
The global parameters $\gamma$ are the slopes that connect the extrinsic-color
parameter and observed magnitudes.  To constrain the degrees of freedom and degeneracies inherent in the model we impose
\begin{equation}
\langle \Delta \rangle=0, \langle k \rangle=0, \langle D \rangle=0, \gamma_0 > 0, \delta_0 < 0.
\end{equation}
The first two specify the definition of zero color relative to which the color excess is measured.    The latter two exclude degenerate posterior space
associated
with  simultaneous sign flips of
$\delta$--$D$ and $\gamma$--$k$.
\color{red}
We assign $k$ and $D$ identical initial conditions, so the difference in the signs of $\gamma_0$ and $\delta_0$ is non-trivial.
Although not formally a degeneracy in the model, practically the parameter pairs  $\delta$--$D$ and $\gamma$--$k$ are degenerate with
each other, as is apparent after 
running the Monte Carlo with exchange of the initial  and boundary conditions of $\delta$ and $\gamma$.
The different initial and boundary conditions for $\gamma_0$ and $\delta_0$ break the degeneracy of our mixture model;
otherwise identical initial conditions for $\gamma$, $\delta$ give them identical posteriors
with broader credible intervals and lower maximum likelihood than when the initial conditions are different.
\color{black}
As the model cannot distinguish which effect is intrinsic or extrinsic, we do so
by assigning the parameters whose values are consistent with dust to $\gamma$ and the others to $\delta$.

@@

\color{red}
The 68\% credible intervals of $\gamma$ are shown in Figure~\ref{fitz:fig}.
A fit of the output $\gamma$ parameters to the dust model of \citet{1999PASP..111...63F} gives $R^F_V=3.06\pm 0.18$.
\color{black}

\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[width=3.2in]{output11/fitz.pdf}
   \caption{\textcolor{red}{68\% credible intervals of $\gamma_X$ and the best-fit to the  \citet{1999PASP..111...63F}  model $R^F_V=3.06$.}
   \label{fitz:fig}}
\end{figure}

%
%The best-fit distributions for $R_X=\frac{\gamma_X}{\gamma_1-\gamma_2}$ are shown in Figure~\ref{rx:fig}.  
%Our measurements of
%$R_X$ and $\frac{\gamma_X}{\gamma_2}$, placed at the effective wavelength of the supernova flux that passes
%through the synthetic bands, are shown in
%Figure~\ref{ccm:fig}.
%
%

%
%\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
%   \centering
%   \includegraphics[width=2.8in]{output11/ccm.pdf}
%      \includegraphics[width=2.8in]{output11/ccm2.pdf} 
%   \caption{68\% credible intervals of Left: $R_X=\frac{\gamma_X}{\gamma_1-\gamma_2}$; Right: $\frac{\gamma_X}{\gamma_2}$. 
%   \label{ccm:fig}}
%\end{figure}

@@

@@

\color{red}
The correlation between $EW_{Si}$ and the SALT2 $X_1$ light-curve shape parameter is verified
for our sample, as shown in Figure~\ref{x1:fig}.
On the other hand, there is no such correlation seen between $E_\delta(B-V)$ and $X_1$.  We conclude that the
intrinsic parameter does not arise to compensate for the fact that light-curve shape is not explicitly
contained in our model.
\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[width=2.8in]{output11/x1si.pdf}
   \includegraphics[width=2.8in]{output11/x1D.pdf}
    \caption{\textcolor{red}{Left: SALT2 $X_1$ versus $EW_{Si}$.  Right: $X_1$ versus $E_\delta(B-V)$.}
   \label{x1:fig}}
\end{figure}

@@

In the fit to our model, the observed color excess is attributed to   extrinsic
$E_\gamma(B-V)=  0.86 \pm   0.08$ and intrinsic $E_\delta(B-V)=  0.35 \pm   0.10$ contributions.
Among the supernovae in the SNfactory  set used to determine $\gamma$ and $\delta$, the
\color{red}
extreme median values for 
$E_\gamma(B-V)$ are $-0.07 \pm 0.01$ and  $  0.33 \pm 0.04$,
and for $E_\delta(B-V)$ $-0.01 \pm 0.01$  and
$  0.07 \pm 0.03$ 
\color{black}
(see Figure~\ref{ebv:fig}).
\textcolor{red}{In both parameters (which are relative to SN~2011fe), SN~2014J lives well outside the 
span} of supernovae used to train the coefficients of the model.

\section{Discussion}
\label{discussion:sec}
We model SNe~Ia broadband optical peak magnitudes allowing for correlations with spectral features at peak and distinct intrinsic and
extrinsic color parameters.  Analyzing SNfactory data with the model, we find significant evidence that the above parameters do
affect supernova magnitudes and colors.  The external color parameter has behavior consistent with $R^F_V=3.06$ \citet{1999PASP..111...63F} dust,
whereas the internal color parameter is inconsistent with dust or a blackbody.  This model  does an excellent job in
describing SN~2014J, a supernova not used in the training that has atypical colors.

\citet{2014ApJ...789...32B, 2015MNRAS.453.3300A} deduce a wide range of dust behavior $1.5<R^F_V<3$ encountered by the SN~Ia population.
\citet{2011ApJ...731..120M, 2011ApJ...729...55F} and the above authors
find that supernovae with large $E_o(B-V)$ 
preferentially exhibit low $R^F_V$. 
\color{red}
In our model internal and external effects for each supernova  conflate into an effective color excess
$E_{eff}(X-V) = E_\gamma(X-V) + E_\delta(X-V)$ and effective $V$ extinction $A_{eff,V} = \gamma_2 k  + \delta_2 D$ 
Recall that the colors where $E_\gamma(B-V)=0$,  $E_\delta(B-V)=0$ are free and arbitrarily
fixed using the  boundary conditions $\langle \Delta \rangle = 0$, $\langle k \rangle = 0$:
there is no unique value for the statistic $R_{eff,V} = A_{eff,V} /E_{eff}(B-V)$, as it depends on the choice of  boundary condition.
For specificity in the current qualitative discussion, 
let us consider a set of SNe~Ia  with parameters set to the median fit values of the SNfactory sample.
We specify   $E_{eff}(U-V)$  and $E_{eff}(B-V)$ by
assigning zero color excess close to the blue-extreme colors of this set, 
as shown in bottom-left corner in the left plot of Figure~\ref{rveff:fig}.
For this choice of color zeropoints,
$R_{eff,U} - R_{eff,V} = E_{eff}(U-V) /E_{eff}(B-V)$ is plotted
in the right of Figure~\ref{rveff:fig}.
Around $E_{eff}(B-V) \sim 0$ the $R_{eff,U} - R_{eff,V}$ values are high, but this is also the regime where their observational  uncertainties 
blow up due to a division of a number close to zero.
The distribution spreads to significantly 
lower   $R_{eff,U} - R_{eff,V}$ as supernovae appear redder but
then reconverge to higher, tighter values red effective color, with asymptote  $R_{eff,U} - R_{eff,V} >1.8$.
This cause of this asymptotic behavior is due to the relatively higher contribution of extrinsic over intrinsic color excess to the effective color excess, as
seen in the bottom plot of
Figure~\ref{rveff:fig}. 
Since the standard deviation of the $E_\delta(B-V)$ distribution is  narrow compared to that of
 $E_\gamma(B-V)$, low values of $R_{eff,U} - R_{eff,V}$ can only occur
at low $E_{eff}(X-V)$.
 The  \citet{1999PASP..111...63F} dust model interprets
 low values of $R_{eff,U} - R_{eff,V}$ as implying high values of $R_V$,
for example $R_{eff,U} - R_{eff,V}=1.70$ for $R_V=3.06$,  $R_{eff,U} - R_{eff,V}=1.81$ for $R_V=1.5$ (see Figure~\ref{ebv:fig}).
The $R_V$ inferred for the reddest supernovae would be $R_V<1.5$, even though their colors are dominated by 
extrinsic $R^F_V=3.06$ dust; this is due our zero colors being (qualitatively) assigned to supernovae that have bluest extrinsic AND intrinsic color excess.
\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[width=2.8in]{output11/rveff.pdf}
   \includegraphics[width=2.8in]{output11/rveff2.pdf}
   \includegraphics[width=2.8in]{output11/rveff3.pdf}
      \caption{\textcolor{red}{Left: Effective color excess $E_{eff}(B-V)$ versus $E_{eff}(U-V)$
   where $E_{eff}(X-V)   = E_\gamma(X-V) + E_\delta(X-V) - z_X$
   for a set of SNe with parameters set to those of the medians of the objects in our analysis.
   The $z_X$ represents a choice of colors at zero color-excess so that the lower left corner points have values approaching zero. 
Right:  The inferred $R_{eff,U}- R_{eff,V} = E_{eff}(U-V)/E_{eff}(B-V)$.
Bottom:  $R_{eff,U}- R_{eff,V}$ as a function of the relative contributions of extrinsic and intrinsic color effects $E_\gamma(B-V)/E_\delta(B-V)$.}
   \label{rveff:fig}}
\end{figure}
\color{black}
